2.3: OLS Linear Regression

### 2019-09-24 static/slides/09-slides.Rmd
Tweak slides 9

- We will begin with .whisper[bivariate] data for relationships between $X$ and $Y$
- Immediate aim is to explore .onfire[associations] between variables, quantified with .shout[correlation] and .shoutcaps[linear regression]
- Later we want to develop more sophisticated tools to argue for .onfire[causation]

@@

```{r, echo = T}
cov(econfreedom$ef,econfreedom$gdp)
```

```{r, echo = T}
econfreedom %>%
  summarize(cov = cov(ef,gdp))
```

@@

- Alternatively, take the average<sup>.red[1]</sup> of the product of standardized ($Z$-scores for) each $(x_i,y_i)$ pair:<sup>.red[2]</sup>

@@

.footnote[<sup>.red[1]</sup> Over `\\(n-1\\)`, since this is a *sample* statistic.

@@

@@

- Your Occasional Reminder: .shoutcaps[Correlation does not imply causation!]
    - I'll show you the difference in a few weeks (when we can actually talk about causation)

@@

@@

- We can make **inferences** about the population slope coefficient
    - eventually, a causal interpretation
    - $\text{slope}=\frac{\Delta Y}{\Delta X}$: for a 1-unit change in $X$, how many units will this *cause* $Y$ to change?

@@

- $\beta_{ClassSize}$ is the .onfire[slope] of the regression line

@@

- So the real question is, .onfire[how can we estimate `\\(\beta_0\\)` and `\\(\beta_1\\)`?]
]

@@

.footnote[<sup>.red[1]</sup> Data are student-teacher-ratio and average test scores on Stanford 9 Achievement Test for 5th grade students for 420 K-6 and K-8 school districts in California in 1999, (Stock and Watson, 2015: p. 141)]
---

@@

$$u = Y_i - \hat{Y_i}$$

@@

### Corplots

```{r}
library(corrplot) # see more at https://github.com/taiyun/corrplot
library(RColorBrewer) # for color scheme used here
M <-cor(mtcars) # need to make a correlation matrix with cor(); can only include numeric variables
corrplot(M, type="upper", 
         method = "circle", # number for showing correlation coefficient
         order="alphabet", 
         col=brewer.pal(n=8, name="RdBu"))
```

# *O*rdinary *L*east *S*quares Estimators

@@

- The solution to the SSE minimization problem yields:.footnote[See **Class Info Page** for proofs.]

@@

>>>>

- We will begin with **bivariate** data for relationships between $X$ and $Y$
- Immediate aim is to explore .onfire[associations] between variables, quantified with .shout[correlation] and .shout[linear regression]
- Later we want to develop more sophisticated tools to argue for .onfire[causation]

@@

```{r, echo = T}
# base R 
cov(econfreedom$ef,econfreedom$gdp)
```

```{r, echo = T}
# dplyr

econfreedom %>%
  summarize(cov = cov(ef,gdp))
```

@@

- Alternatively, take the average<sup>.red[1]</sup> of the product of standardized $(Z$-scores for) each $(x_i,y_i)$ pair:<sup>.red[2]</sup>

@@

.footnote[<sup>.red[1]</sup> Over `\\((n-1)\\)`, since this is a *sample* statistic.

@@

# Correlation and Covariance in R

```{r, echo = T}
# Base r: cov or cor(df$x, df$y)

cov(econfreedom$ef, econfreedom$gdp)

cor(econfreedom$ef, econfreedom$gdp)

# dplyr

econfreedom %>%
  summarize(covariance = cov(ef, gdp),
            correlation = cor(ef, gdp))

```
---

# Correlation and Covariance in R I

- `corrplot` is a great package (install and then load) to **visualize** correlations in data

```{r, echo=T, eval=F}
library(corrplot) # see more at https://github.com/taiyun/corrplot
library(RColorBrewer) # for color scheme used here
library(gapminder) # for gapminder data

# need to make a corelation matrix with cor(); can only include numeric variables
gapminder_cor<- gapminder %>%
  dplyr::select(gdpPercap, pop, lifeExp)

# make a correlation table with cor (base R)
gapminder_cor_table<-cor(gapminder_cor)

# view it
gapminder_cor_table
```

# Correlation and Covariance in R II

```{r, echo=T, eval=F}

corrplot(gapminder_cor_table, type="upper", 
         method = "circle", # number for showing correlation coefficient
         order="alphabet", 
         col=brewer.pal(n=8, name="RdBu"))
]

```{r, fig.retina=3}
library(corrplot) # see more at https://github.com/taiyun/corrplot
library(RColorBrewer) # for color scheme used here
library(gapminder) # for gapminder data

# need to make a corelation matrix with cor(); can only include numeric variables
gapminder_cor<- gapminder %>%
  dplyr::select(gdpPercap, pop, lifeExp)

# make a correlation table with cor (base R)
gapminder_cor_table<-cor(gapminder_cor)

corrplot(gapminder_cor_table, type="upper", 
         method = "circle", # number for showing correlation coefficient
         order="alphabet", 
         col=brewer.pal(n=8, name="RdBu"))

@@

- Your Occasional Reminder: .shout[Correlation does not imply causation!]
    - I'll show you the difference in a few weeks (when we can actually talk about causation)

@@

.center[
![](https://www.dropbox.com/s/v5vwsadw5vs448t/causality.jpg?raw=1)
]
]

@@

- We can make **inferences** about the population slope coefficient

- eventually, a causal interpretation

- $\text{slope}=\frac{\Delta Y}{\Delta X}$: for a 1-unit change in $X$, how many units will this *cause* $Y$ to change?

@@

- $\beta_{1}$ is the .onfire[slope] of the regression line

@@

- So the real question is, .onfire[how can we estimate `\\(\beta_0\\)` and `\\(\beta_1\\)`?]

.footnote[<sup>.red[1]</sup> Data are student-teacher-ratio and average test scores on Stanford 9 Achievement Test for 5th grade students for 420 K-6 and K-8 school districts in California in 1999, (Stock and Watson, 2015: p. 141)]

@@

@@

@@

# *O* rdinary *L* east *S* quares Estimators

@@

- The solution to the SSE minimization problem yields:<sup>.red[1]</sup>

@@

.footnote[<sup>.red[1]</sup> See [tomorrow's class notes page](/class/10-class) for proofs.]
---
